Evaluation of Diagnostic Concordance Between Algorithms for Parkinson’s Disease Dementia

### 2025-10-09 _manuscript/manuscript.qmd
Aligning the quarto version with current manuscript on GDrive for easier manipulation during review.

upstore <- here::here(tar_config_get("store"))
tar_source(here::here("R"))
data <- tar_read(raw_data, store = upstore)
crit <- tar_read(pdd_data, store = upstore)$criteria
vars <- tar_read(variables, store = upstore)
```

@@

According to a recent meta-analysis, approximately one-quarter of PD patients is likely to be diagnosed with PDD [@sousa2022]. However, reported PDD rate estimates vary widely, ranging from 14% up to 55%, depending on methodological criteria employed [@sousa2022]. Moreover, factors such as patients' sex [@cereda2016], age and disease duration appear to modulate the risk of cognitive decline and PDD [@oh2016; @rana2012].

Despite the clinical relevance of PDD, its diagnosis remains complex. A milestone in research of PDD was the publication of diagnostic criteria established in 2007 by the International Parkinson and Movement Disorder Society (MDS) [@dubois2007]. In these citeria, the MDS introduced a two-levelled system for PDD detection. Level I consists of brief cognitive assessments, while Level II involves comprehensive neuropsychological testing across cognitive domains [@emre2007].

The original Level I algorithm included eight conditions that had to be satisfied simultaneously in order to diagnose probable PDD. These included: 1) diagnosis of PD proposed by the Queen Square Brain Bank; 2) PD onset prior to the PDD emergence; 3) evidence of global cognitive impairment (MMSE score \< 26 points); 4) cognitive deficit interference with the IADL (assessed by the pill questionnaire or caregiver interview); 5) impairment in at least two cognitive domains, namely memory, attention, visuo-constructive abilities and executive functions; 6) there was absence of Major Depressive Disorder; 7) absence of delirium; and 8) exclusion of other abnormalities and potential causes of dementia [@dubois2007].

Currently, efforts are focused on refining this PDD diagnostic framework. A recent call for a change pinpoints limitations regarding the original criteria and suggest various updates to enhance their utility [@kulisevsky2024]. Proposed suggestions include replacement of Mini Mental State Examination (MMSE) by Montreal Cognitive Assessment (MoCA), which is more sensitive to PD specific cognitive impairment; expansion of instrumental activities of daily living (IADL) evaluation; inclusion of language assessment; recognition of anxiety as one the neuropsychiatric symptoms relevant in PDD; and intergration of biomarkers.

@@

Cognitive performance was evaluated at both Level I and Level II according to the standard MDS battery for Parkinson's Disease Mild Cognitive Impairment (PD-MCI) [@litvan2012, @bezdicek2017]. Cognitive performance at Level I was assessed by the Mini-Mental State Examination (MMSE) [@stepankova2014; @folstein1975] and the Montreal Cognitive Assessment (MoCA) [@kopecek2016; @nasreddine2005]. The comprehensive neuropsychological assessment at Level II evaluated five cognitive domains through specific test: attention and working memory assessed by Trail Making Test Part A (TMT-A) [@bezdicek2012; @reitan2004], and WAIS Digit Span Backward (WAIS DSB) [@wechsler1997], executive functions by Categorical Verbal Fluency (CF) [@nikolai2015], and subtest from the Prague Stroop Test – Colors (PST-C) [@bezdicek2021], language by the WAIS Similarities subtest [@wechsler1997], and the Boston Naming Test (BNT-60) [@zemanova2016; @kaplan1983], memory by the Rey Auditory Verbal Learning Test (RAVLT) [@frydrychova2018; @bezdicek2013; @rey1964] delayed recall, and the Brief Visual Memory Test–Revised (BVMTR) [@havlik2020; @benedict1997] delayed recall, or WAIS Family Pictures subtest [@wechsler1997] delayed recall, visuospatial function assessed by the Judgment of Line Orientation Test (JoLO) [@benton1983], and Clock Drawing Test (CLOX) [@royall1998].

The Functional Activities Questionnaire (FAQ) [@bezdicek2016; @pfeffer1982] was administered to assess functional impairment. The Beck Depression Inventory-II (BDI-II) [@ciharova2020; @beck1996] and State-Trait Anxiety Inventory (STAI) [@spielberg1983; @mullner1980] were used to assess neuropsychiatric status. Psychotic symptoms were assessed through structured psychiatric interviews conducted by a trained psychiatrist.

@@

In this study, we applied three distinct sets of diagnostic algorithms for probable PDD at Level I. The first set was based on the original framework [@dubois2007], which utilized the Mini-Mental State Examination (MMSE) as a global cognitive screening tool, supplemented by assessments of attention, executive function, visuospatial abilities, and memory. The second set of algorithms was based on the recent call for change of dementia diagnostic guidelines [@kulisevsky2024], which advocates for more sensitive cognitive domain assessments in the context of PD. This updated approach incorporated specific items from the Montreal Cognitive Assessment (MoCA). The third approach applied the Czech version of the shortened Montreal Cognitive Assessment (sMoCA) [@bezdicek2020], a time-efficient modification designed to measure global cognitive performance using a reduced testing protocol that omits items providing redundant information. Lastly, the fourth approach followed the Level II protocol for diagnosis of PDD and Mild Cognitive Impairment in PD (PD-MCI) [@dubois2007; @litvan2012]. The Level II methodology, including the use of a regression-based normative scoring approach, has been detailed in a prior study [@bezdicek2017]. In this study, the thresholds for cognitive impairment at Level II were set at $z \leq -1.5$. All non-cognitive criteria of probable PDD (i.e., diagnosis of PD that developed before dementia and absence of Major Depression, delirium or other abnormalities that obscure diagnosis) held true for all patients in the sample according to the psychiatric and neurological examinations.

@@

Following the framework proposed by @lundberg2021, in this study we explicitly connect our research objectives and their corresponding theoretical (i.e., targets of inference) and empirical (i.e., data-driven) estimands to statistical estimates. The theoretical estimand refers to a unit-specific quantity defined over a target population and represents the ideal quantity that would address the research question under optimal conditions, such as access to complete population data or perfect experimental control. In contrast, the empirical estimand corresponds to the quantity that is actually computable using the available dataset, given real world constraints. The full description of the study's estimands and their relation to our research objectives is presented in the Appendix (see @tbl-estimands).

To address study objectives, we started by repeatedly assigning each patient the diagnosis of probable PDD based on each PDD algorithm listed in @tbl-crits (see also @tbl-algos) resulting in a `r sum(data$incl)` (patients) $\times$ `r sum(critnum)` (algorithms) matrix where each cell indicates whether a patient (row) meets criteria for probable PDD according to an algorithm (column). PDD rate estimates were computed as $\frac{N_{PDD}}{N_{total}}$ separately for each algortihm. The predictive value of age and sex was then evaluated by fitting a set of logistic regressions, one for each algorithm for probable PDD, whereby the probable PDD was predicted by age, sex and their interaction.

@@

# In-text demographic data
sex <- subset(desc$table, variable == "Sex")$N
age <- paste0(subset(desc$table, variable == "Age")[ , c("M", "SD")] |> paste(collapse = " (SD = "),") years of age")
edu <- paste0(subset(desc$table, variable == "Education")[ , c("M", "SD")] |> paste(collapse = " (SD = ")," ) years of education")
dur <- paste0(subset(desc$table, variable == "PD duration")[ , c("M", "SD")] |> paste(collapse = " (SD = "),") years of disease duration")
updrs_off <- paste0(subset(desc$table, variable == "UPDRS III off state")[ , c("M", "SD")] |> paste(collapse = " (SD = "),") Unified Parkinson Disease Rating Scale (UPDRS), part III in medication OFF state")
updrs_on <- paste0(subset(desc$table, variable == "UPDRS III on state")[ , c("M", "SD")] |> paste(collapse = " (SD = "),") UPDRS III in medication ON state")

@@

A total of `r sum(data$incl)` patients were included. The sample included `r sex` men, with an average of `r age`, `r edu`, `r dur`, `r updrs_off` and `r updrs_on`. Cognitive characteristics of the sample are summarized in @tbl-desc.

@@

Interestingly, we did not observe any reliable age-related differences in PDD rate within our cohort. This lack of age-dependency may, however, be also stemming from the relatively younger age of our cohort, because previous reports indicate that association between age and PDD is not linear but increases with age and may not reach substantial values before the age of 70 [@rana2012].

@@

The advantage of MoCA-based algorithm was evident in combination with using the FAQ total score, reaching accuracy rates of 94–99%. Furthermore, in the algorithm using sMoCA, the raw accuracy was moderate, however, the balanced accuracy (i.e. combined sensitivity and specificity) was high. These findings imply that simplified tools such as sMoCA may be a promising tool for heightening efficiency of the assessment while maintaining diagnostic accuracy.

@@

Nevertheless, the analysis of the algorithm concordance might not be affected by the sample to the same degree allowing for a broader generalization. Future studies should aim to replicate our results on larger and different cohorts of PD patients. To make this process easier, the code used to generate our results is publicly available and easily applicable to similarly structured data.

Furthermore, another limitation is the use of the FAQ questionnaire for IADL assessment. FAQ is subjective and informant-reliant, which may result in biased data. For future research, we suggest the use of more objective, modernized tools for IADl assessment, such as performance-based methods [@schmitter-edgecombe2020] or questionnaire adaptation including questions regarding gadgets use and digital literacy [@postema2024]. Our results emphasize the importance of IADL deficit for the PDD diagnosis, therefore, we suggest exploring more reliable tools with high ecological validity.

Finally, whereas our study systematically investigated how varying definition of global deficit, impaired cognition and IADL deficit affect probable PDD classification, it did not explore associations of PDD diagnosis with its neuropsychiatric (e.g. anxiety profile) and biomarker correlates. Future studies would benefit from focusing on these aspects of PDD.

@@

[^3]: Because the quantity of interest is a rate and could thus be though of as a sum of binomially distributed PDD occurences divided by the total number of patients, its variance will likely systematically vary with its mean. Specifically, as the rate goes from extremes to 0.5, the variance increases. Consequently, if our estimate of the rate was lower than the true population rate, e.g., because our sample includes younger patients compared to the general PD population, our estimate of variance would also lower than the true variance of PDD rate in the general PD population. Nonetheless, the between-algorithm variability may not be affected by this phenomenon as unlike variability of PDD rate, we do not have reason to assume it comes about by summing independent binomial events

>>>>

upstore <- here::here(tar_config_get("store"))
tar_source(here::here("R"))

data <- tar_read(raw_data, store = upstore)
crit <- tar_read(pdd_data, store = upstore)$criteria
vars <- tar_read(variables, store = upstore)
```

@@

According to a recent meta-analysis, approximately one-quarter of PD patients is likely to be diagnosed with PDD [@sousa2022]. However, reported PDD rate estimates vary widely, ranging from 14% up to 55%, depending on the methodological criteria employed [@sousa2022]. Moreover, factors such as patients' sex [@cereda2016], age and disease duration appear to modulate the risk of cognitive decline and PDD [@oh2016; @rana2012].

Despite the clinical relevance of PDD, its diagnosis remains complex. A milestone in research of PDD was the publication of diagnostic criteria established in 2007 by the International Parkinson and Movement Disorder Society (MDS) [@dubois2007]. In these criteria, the MDS introduced a two-levelled system for PDD detection. Level I consists of brief cognitive assessments, while Level II involves comprehensive neuropsychological testing across cognitive domains [@emre2007].

The original Level I algorithm included eight conditions that had to be satisfied simultaneously in order to diagnose probable PDD. These included: 1) diagnosis of PD proposed by the Queen Square Brain Bank; 2) PD onset prior to the PDD emergence; 3) evidence of global cognitive impairment (MMSE score \< 26 points); 4) cognitive deficit interference with the IADL (assessed by the pill questionnaire or caregiver interview); 5) impairment in at least two cognitive domains, namely memory, attention, visuo-constructive abilities and executive function; 6) there was absence of Major Depressive Disorder; 7) absence of delirium; and 8) exclusion of other abnormalities and potential causes of dementia [@dubois2007].

Currently, efforts are focused on refining this PDD diagnostic framework. A recent call for a change pinpoints limitations regarding the original criteria and suggest various updates to enhance their utility [@kulisevsky2024]. Proposed suggestions include replacement of Mini Mental State Examination (MMSE) by Montreal Cognitive Assessment (MoCA), which is more sensitive to PD specific cognitive impairment; expansion of instrumental activities of daily living (IADL) evaluation; inclusion of language assessment; recognition of anxiety as one **of** the neuropsychiatric symptoms relevant in PDD; and intergration of biomarkers.

@@

Cognitive performance was evaluated at both Level I and Level II according to the standard MDS battery for Parkinson’s Disease Mild Cognitive Impairment (PD-MCI) [@litvan2012, @bezdicek2017]. Cognitive performance at Level I was assessed by the Mini-Mental State Examination (MMSE) [@stepankova2014; @folstein1975] and the Montreal Cognitive Assessment (MoCA) [@kopecek2016; @nasreddine2005]. The comprehensive neuropsychological assessment at Level II evaluated five cognitive domains through specific tests: attention and working memory assessed by Trail Making Test Part A (TMT-A) [@bezdicek2012; @reitan2004], and WAIS Digit Span Backward (WAIS DSB) [@wechsler1997], executive function by Categorical Verbal Fluency (CF) [@nikolai2015], and subtest from the Prague Stroop Test – Colors (PST-C) [@bezdicek2021], language by the WAIS Similarities subtest [@wechsler1997], and the Boston Naming Test (BNT-60) [@zemanova2016; @kaplan1983], memory by the Rey Auditory Verbal Learning Test (RAVLT) [@frydrychova2018; @bezdicek2013; @rey1964] delayed recall, and the Brief Visual Memory Test–Revised (BVMTR) [@havlik2020; @benedict1997] delayed recall, or WAIS Family Pictures subtest [@wechsler1997] delayed recall, visuospatial function assessed by the Judgment of Line Orientation Test (JoLO) [@benton1983], and Clock Drawing Test (CLOX) [@royall1998].

The Functional Activities Questionnaire (FAQ) [@bezdicek2016; @pfeffer1982] was administered to assess functional impairment. The Beck Depression Inventory-II (BDI-II) [@ciharova2020; @beck1996] and State-Trait Anxiety Inventory (STAI) [@spielberg1983; @mullner1980] were used to assess neuropsychiatric status. **BDI-II and STAI are a part of the standard methods in our pre-DBS protocol, having well-established psychometric properties in PD populations [@leentjens2000; @dissanayaka2010]. Other psychiatric scales data, such as the Neuropsychiatric Inventory, were not available for our retrospective dataset. However, all patients underwent a comprehensive evaluation by a trained neuropsychiatrist, excluding the presence of major depression, apathy, psychosis, and visual hallucinations, ensuring that confounding neuropsychiatric conditions were not present in our sample.**

@@

In this study, we applied three distinct sets of diagnostic algorithms for probable PDD at Level I. The first set was based on the original framework [@dubois2007], which utilized the Mini-Mental State Examination (MMSE) as a global cognitive screening tool, supplemented by assessments of attention, executive function, visuospatial abilities, and memory. The second set of algorithms was based on the recent call for change of dementia diagnostic guidelines [@kulisevsky2024], which advocates for more sensitive cognitive domain assessments in the context of PD. This updated approach incorporated specific items from the Montreal Cognitive Assessment (MoCA). The third approach applied the Czech version of the shortened Montreal Cognitive Assessment (sMoCA) [@bezdicek2020], a time-efficient modification designed to measure global cognitive performance using a reduced testing protocol that omits items providing redundant information. **The sMoCA has been validated in Czech PD cohorts [@bezdicek2020] and shown to be sensitive to cognitive deficits while lowering patient burden [@roalf2017a]. Inclusion of the sMoCA in our study was motivated by its clinical utility in pre-surgical settings, where time restrictions and patients’ fatigue often limit the feasibility of longer assessments. Moreover, its inclusion allowed us to compare diagnostic performance across different screening lengths.**

Lastly, the fourth approach followed the Level II protocol for diagnosis of PDD and Mild Cognitive Impairment in PD (PD-MCI) [@dubois2007; @litvan2012]. The Level II methodology, including the use of a regression-based normative scoring approach, has been detailed in a prior study [@bezdicek2017]. In this study, the thresholds for cognitive impairment at Level II were set at $z \leq -1.5$. All non-cognitive criteria of probable PDD (i.e., diagnosis of PD that developed before dementia and absence of Major Depression, delirium or other abnormalities that obscure diagnosis) held true for all patients in the sample according to the psychiatric and neurological examinations.

@@

Following the framework proposed @lundberg2021, in this study we explicitly connect our research objectives and their corresponding theoretical (i.e., targets of inference) and empirical (i.e., data-driven) estimands to statistical estimates. The theoretical estimand refers to a unit-specific quantity defined over a target population and represents the ideal quantity that would address the research question under optimal conditions, such as access to complete population data or perfect experimental control. In contrast, the empirical estimand corresponds to the quantity that is actually computable using the available dataset, given real world constraints. The full description of the study’s estimands and their relation to our research objectives is presented in the Appendix (see @tbl-estimands).

To address study objectives, we started by repeatedly assigning each patient the diagnosis of probable PDD based on each PDD algorithm listed in @tbl-crits (see also @tbl-algos) resulting in a `r sum(data$incl)` (patients) $\times$ `r sum(critnum)` (algorithms) matrix where each cell indicates whether a patient (row) meets criteria for probable PDD according to an algorithm (column). PDD rate estimates were computed as $\frac{N_{PDD}}{N_{total}}$ separately for each algorithm. The predictive value of age and sex was then evaluated by fitting a set of logistic regressions, one for each algorithm for probable PDD, whereby the probable PDD was predicted by age, sex and their interaction.

@@

# Function for writing M (SD):
msd_txt <- function(x, txt = "", tab = desc$table) {
  paste0(
    subset(tab, variable == x)[ , c("M", "SD")] |> paste(collapse = " (SD = "),
    ")",
    txt
  )
}

# In-text demographic data
sex <- subset(desc$table, variable == "Sex")$N
age <- msd_txt("Age", " years of age")
edu <- msd_txt("Education", " years of education")
dur <- msd_txt("PD duration", " years of disease duration")
updrs_off <- msd_txt("UPDRS III off state", " Unified Parkinson Disease Rating Scale (UPDRS), part III in medication OFF state")
updrs_on <- msd_txt("UPDRS III on state", " Unified Parkinson Disease Rating Scale (UPDRS), part III in medication ON state")
bdi <- msd_txt("BDI")
staix1 <- msd_txt("STAI X1")
staix2 <- msd_txt("STAI X2")

@@

A total of `r sum(data$incl)` patients were included. The sample included `r sex` men, with an average of `r age`, `r edu`, `r dur`, `r updrs_off` and `r updrs_on`. **Descriptive statistics for neuropsychiatric symptoms indicated mild depressive and anxiety symptoms in our cohort, with the average BDI-II of `r bdi`, average STAI X1 of `r staix1`, and average STAI X2 of `r staix2`. However, according to the psychiatric assessment, none of the patients with probable PDD was suffering major depressive disorder, delirium or other neuropsychiatric abnormalities that would disqualify the diagnosis.** Cognitive characteristics of the sample are summarized in @tbl-desc.

@@

Interestingly, we did not observe any reliable age-related differences in PDD rate within our cohort. This lack of age-dependency may, however, be also stemming from the relatively younger age of our cohort, because previous reports indicate that association between age and PDD is not linear but increases with age and may not reach substantial values before the age of 70 [@rana2012]. **In our sample, the younger age and preserved cognitive level might limit age-related effects. This fact could bias algorithm results toward higher specificity but lower sensitivity [@larvin2023]. Previous studies report that DBS candidates tend to be, on average, younger and cognitively less impaired than the general PD population [@bronstein2011], which could bias diagnostic algorithm performance toward overestimating specificity. Therefore, our findings may not be fully generalizable to older or more cognitively impaired people with PD.**

@@

The advantage of MoCA-based algorithm was evident in combination with using the FAQ total score, reaching accuracy rates of 94–99%. Furthermore, in the algorithm using sMoCA, the raw accuracy was moderate, however, the balanced accuracy (i.e. combined sensitivity and specificity) was high. These findings imply that simplified tools such as sMoCA may be a promising tool for heightening efficiency of the assessment while maintaining diagnostic accuracy. **This underscores the clinical utility of sMoCA screening tool. Although raw accuracy was moderate, balanced accuracy results ensure avoidance of bias from class imbalance. From a practical point of view, it suggests that even brief screening tools can achieve reliable classification, which means an advantage for routine clinical settings and time-restricted assessments.**

@@

Nevertheless, the analysis of the algorithm concordance might not be affected by the sample to the same degree allowing for a broader generalization. **However, the use of a DBS cohort offered certain methodological advantages. For example, all patients underwent standardized and comprehensive neuropsychological testing, which resulted in well-characterized and high-quality dataset, allowing us to systematically evaluate multiple diagnostic algorithms.** Future studies should aim to replicate our results on larger and different cohorts **with more heterogenous sample**. To make this process easier, the code used to generate our results is publicly available and easily applicable to similarly structured data.

Furthermore, another limitation is the use of the FAQ questionnaire for IADL assessment. FAQ is subjective and informant-reliant, which may result in biased data. **At the same time, the questionnaire may vary across cultures, therefore, the results may not be transferable. For example, activities such as financial management or driving, may not be equally practiced across cultures or contexts, which might complicate interpretation of impairment levels. The FAQ scores may vary across cultures, and diagnostic thresholds may not be transferable across countries [@odonald2025]. This may influence both the sensitivity and ecological validity of the functional criteria used. Another limitation is that the FAQ scores may correlate with neuropsychological results of certain domains, specifically executive functions or memory [@vanrentergem2020; @moheb2017]. This overlap could artificially increase concordance between cognitive and functional criteria. For future research, we suggest using a PD-specific questionnaire, such as the Penn Parkinson's Daily Activities Questionnaire-15 [@brennan2016] or generally more objective methods, preferably performance-based ones. Our results underscore the importance of IADL measurement for the PDD diagnosis. Therefore, we recommend using more reliable tools with high ecological validity.** For future research, we suggest the use of more objective, modernized tools for IADl assessment, such as performance-based methods [@schmitter-edgecombe2020] or questionnaire adaptation including questions regarding gadgets use and digital literacy [@postema2024]. Our results emphasize the importance of IADL deficit for the PDD diagnosis, therefore, we suggest exploring more reliable tools with high ecological validity.

**Additionally, the current study did not explore associations of PDD diagnosis with its neuropsychiatric correlates and biomarkers. We recommend examining models where psychiatric symptoms and biomarkers are treated as covariates to better capture their impact on PDD diagnosis.**

??Finally, whereas our study systematically investigated how varying definition of global deficit, impaired cognition and IADL deficit affect probable PDD classification, it did not explore associations of PDD diagnosis with its neuropsychiatric (e.g. anxiety profile) and biomarker correlates. Future studies would benefit from focusing on these aspects of PDD.??^[**JM: Why was the previous paragraph added instead of this one? This looks perfectly functional. The same goes largely for the IADL discussion, the original was better (in terms of resources and logical coherence) at most places.**]

@@

[^3]: Because the quantity of interest is a rate and could thus be though of as a sum of binomially distributed PDD occurences divided by the total number of patients, its variance will likely systematically vary with its mean. Specifically, as the rate goes from extremes to 0.5, the variance increases. Consequently, if our estimate of the rate was lower than the true population rate, e.g., because our sample includes younger patients compared to the general PD population, our estimate of variance would also be lower than the true variance of PDD rate in the general PD population. Nonetheless, the between-algorithm variability may not be affected by this phenomenon as unlike variability of PDD rate, we do not have reason to assume it comes about by summing independent binomial events.
